---
title: "IS 607 Project 2"
author: "Derek G. Nokes"
date: "Saturday, March 14, 2015"
output: html_document
toc: yes
---

\pagebreak

```{r,warning=FALSE,error=FALSE,message=FALSE}
# load the libraries
library('ggplot2')
library('ggthemes')
library('RColorBrewer')
library('Hmisc')
```

## Load Data

We begin by loading the data.

```{r}
# load the data
inputFile<-"C:/Users/dgn2/Documents/R/IS607/Project_2/project2_data.csv"
# load data
data <- read.csv(inputFile, header=TRUE)
```

The data is displayed in the following table:

```{r,echo=FALSE}
# create the indices
indexI<-data[,1]=='I'
indexII<-data[,1]=='II'
indexIII<-data[,1]=='III'
indexIV<-data[,1]=='IV'
# create the data table
tableData<-data.frame(data[indexI,2:3],data[indexII,2:3],
                      data[indexIII,2:3],data[indexIV,2:3])
# create the column labels
dataLabels<-c('I X','I Y','II X','II Y','III X','III Y',
              'IV X','IV Y')
# label the data
colnames(tableData)<-dataLabels
# create the table
knitr::kable(tableData, caption = 'X and Y by Quarter')
```

## Exploratory Data Analysis

### Data Description
Once the data is loaded, we take a quick look at the attributes of the data:

```{r}
str(data)
```

```{r,echo=FALSE}
# find the shape of the data
dimension<-dim(data)
nRows<-dimension[1]
nCols<-dimension[2]
```

We can see that there are `r nRows` rows and `r nCols` columns in our data 
set (i.e., $quarter$, $x$, and $y$).

```{r,echo=FALSE}
# create the data table
dataDescription<-data.frame(c('quarter','x','y'),
                            c('is a factor with 4 levels (I, II, III, & IV)',
                              'is an int','is a decimal number'))
# label the data
colnames(dataDescription)<-c('variable','description')
# create the table
knitr::kable(dataDescription, caption = 'Data Description')
```

A quick summary of the data by the $quarter$ factor does not provide much insight.

```{r}
# create the data summary
summaries <- aggregate(data$y, by=list(data$quarter),FUN=summary)
# relabel the data summary
colnames(summaries)<-c('Group','')
summaries
```

### Graphical Exploration

First, we create a single scatter plot varying the color by quarter.

```{r}
# graph the data
p <- ggplot(data, aes(x, y))+theme_economist() + 
  scale_colour_economist() + ggtitle("Scatterplots of X vs. Y") + 
  geom_rug()
p + geom_point(aes(colour = quarter))
```

The patterns associated with each quarter are difficult to see 
on a single scatter plot, so we split out the quarters into 
separate scatter plots.

```{r}
qplot(x,y, data=data,color=quarter,facets=~quarter,
      xlab="x", ylab="y",main="Scatterplots of X vs. Y") + 
  geom_rug() + scale_fill_brewer() + 
  theme_economist() + scale_colour_economist() 
```

The patterns in each data set are clearer. The I, III and IV data appear
to have linear relationships. The II data is clearly not linear. Both III 
and IV appear to have outliers that obscure the data. 

```{r}
p <- ggplot(data, aes(quarter,y))
p + geom_boxplot(outlier.colour = "blue",aes(fill = quarter)) + 
  coord_flip() + scale_fill_brewer() + ggtitle('Y Summary By Quarter')+
  theme_economist() + scale_colour_economist()
```

We now examine the distribution of Y, first by creating a 
histogram where color denotes the contribution by each quarter

```{r}
# create a histogram where color denotes contribution by
# each quarter
dplot <- ggplot(data, aes(y, fill = quarter))
dplot + geom_bar(position = "stack",binwidth=1) + 
  theme_economist() + scale_colour_economist() + 
  scale_fill_brewer() + ggtitle('Y By Quarter')
```

We continue to examine the distribution of Y, focusing next 
on kernel smoothed densities for each respective quarter.

```{r}
# plot the kernal smoothed density by quarter
qplot(y, data=data, geom="density",position="stack", fill=quarter,
      alpha=I(.6), main="Distribution of Y BY Quarter",
      xlab="y", ylab="density") + scale_fill_brewer() + 
  theme_economist() + scale_colour_economist() 
```

The shape of the distribution of Y for each quarter is more clear
once we look at the distribution for each quarter separately.

```{r}
# plot the kernal smoothed density by quarter
qplot(y, data=data, geom="density",position="stack", fill=quarter,
      alpha=I(.6), main="Distribution of Y BY Quarter",
      xlab="y", ylab="density") + scale_fill_brewer() + 
  theme_economist() + scale_colour_economist() + 
  facet_grid(. ~ quarter) +  coord_flip()
```

The density differs significantly by quarter. The mode of the 
distribution shifts.